---
title: "ENV797_TSA_project_S2024"
author: "Faustin Kambale Samantha Pace Drew Wolanski"
date: "2024-03-22"
output: pdf_document
editor_options: 
  chunk_output_type: console
---


```{r, message=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(cowplot)
library(sarima)
library(patchwork)
library(dplyr)
library(forecast)
library(readxl)
library(kableExtra)
```

## CO2 Emissions Forecasting : Predicting U.S. CO2 emissions from energy consumption

# Background & Research Question

1. Coal has been on the decline due to poor economics and increased scrutiny on emissions
2. Natural gas has concurrently been on the rise due to low prices caused by fracking revolution
   - Decrease in coal emissions partially offset by new natural gas plants
3. Project aims to forecast the CO2 emissions from coal, natural gas, and petroleum based on
historical data and determine if total CO2 emissions are declining as needed to meet climate goals
  - Biden goal to reduce GHG emissions by 50% below 2005 levels by 2030

# Data description

To verify our research question, we used data from the U.S. Energy Information Administration *Monthly Energy Review February 2024*

Data Category: Environment
Energy Sources: Coal, Natural Gas, Petroleum
Sampling frequency: Monthly
Training Data Date Range: Jan 1973-Nov 2023 (n=611)
Forecast Length: 6 years (2030)

## Loading the Dataset

```{r}
data <- read_excel ("/Users/faustinkambale/Library/CloudStorage/OneDrive-DukeUniversity/Spring 2024 classes/Time Series 4 Energy/TSA_Sp24/Project/ENV797_TSA_project_S2024/Data/Table_11.1_Carbon_Dioxide_Emissions_From_Energy_Consumption_by_Source.xlsx",  skip = 10)
```

##Data wrangling: remaning the variables

```{r}
#Rename the columns 
data <- data %>%
  rename (
    coalimports = `Coal, Including Coal Coke Net Imports, CO2 Emissions (Million Metric Tons of Carbon Dioxide)`, 
    naturalgas = `Natural Gas, Excluding Supplemental Gaseous Fuels, CO2 Emissions (Million Metric Tons of Carbon Dioxide)`,
    fuleoil= `Distillate Fuel Oil, Excluding Biodiesel, CO2 Emissions (Million Metric Tons of Carbon Dioxide)`,
    aviationgas = `Aviation Gasoline CO2 Emissions (Million Metric Tons of Carbon Dioxide)`,
    hydrocabongas= `Hydrocarbon Gas Liquids CO2 Emissions (Million Metric Tons of Carbon Dioxide)`,
    jetco2emmissions= `Jet Fuel CO2 Emissions (Million Metric Tons of Carbon Dioxide)`,
    keroseneco2emissions=`Kerosene CO2 Emissions (Million Metric Tons of Carbon Dioxide)`,
    lubricantsco2emissions=`Lubricants CO2 Emissions (Million Metric Tons of Carbon Dioxide)`,
    motosgasolineco2emissions=`Motor Gasoline, Excluding Ethanol, CO2 Emissions (Million Metric Tons of Carbon Dioxide)`,
    petroliumco2emissions=`Petroleum Coke CO2 Emissions (Million Metric Tons of Carbon Dioxide)`,
    residualco2emissions=`Residual Fuel Oil CO2 Emissions (Million Metric Tons of Carbon Dioxide)`,
    otherpetroliumco2emissions=`Other Petroleum Products CO2 Emissions (Million Metric Tons of Carbon Dioxide)`,
    petroliumlessbioco2emissions=`Petroleum, Excluding Biofuels, CO2 Emissions (Million Metric Tons of Carbon Dioxide)`,
    totenergyco2emissions=`Total Energy CO2 Emissions (Million Metric Tons of Carbon Dioxide)`
    )

#Inspecting the Data
data$Month <- as.Date(data$Month)

#Setting Data to work with
ourdata <- as.data.frame(data[,c("Month","coalimports", "naturalgas", "petroliumlessbioco2emissions")])
str(ourdata)
head(ourdata, 10)
tail(ourdata, 10)
summary(ourdata, 10)

coal_ts <- ts(ourdata[,2], start=c(1973,1),frequency=12)
plot(coal_ts)
```

##Data analysis 

# Ploting ACF and PACF

```{r}
coal_acf <- Acf(coal_ts, lag.max = 40, main = "Co2 Emission from Coal Imports")
coal_pacf <-Pacf(coal_ts, lag.max = 40, main = "Coal Imports")
plot_grid (
  autoplot(coal_acf, main = "Coal Imports"),
  autoplot(coal_pacf, main = "Coal Imports")
)
```

Reading the ACF we clearly notice a gradual decline and positive autororrelation in our Coal import data set. This pattern suggest that the data is non-stationary. The PACF, with a cut off after lag 1. Visual look suggest the presence of seasonal effect in our data set.The Pac anf Acf plots suggest that a AR (1) model can best fit the data.  

# Decomposing Data

```{r}
coal_decomp <- mstl(ts_data[,2])
autoplot(coal_decomp, main = "CO2 Emmission from Coal Imports")
```
Decomposing the data set gives a clear picture of the decreasing trend in the coal import data and the presence of the seasonality in our data set. To check a little deeper, we zoom in with a small time period. 

#Zooming in a small data sample can be ignored

```{r}
data_filtered <- filter(ourdata, year(Month) >= 2010)
coal_ts1 <- ts(data_filtered[,2], start=c(2010,1),frequency=12)
plot(coal_ts1)
coal_decomp1 <- plot(decompose(coal_ts1, type = "additive"))
```

As we could have noticed before, this subet of the coal data show the presence of a seasonality component.

##Forecasting 
For a better forecast, we subset our data into train and test data for a horizon of 85 months. The Train data set serves to test the best fit based on the historical data whereas the test data set serves for testing the forecast. 
```{r message=FALSE, warning=FALSE}
#create a subset data for training
n_for = 85
coal_ts_train <- subset(coal_ts,end = length(coal_ts)-n_for)
#create a subset data for experiment
coal_ts_test <- subset(coal_ts, start = length(ts_faustin)-n_for+1)
print(start(coal_ts_train))
print(end(coal_ts_train))
print(start(coal_ts_test))
print(end(coal_ts_test))

autoplot(coal_ts_train)
autoplot(coal_ts_test)
```

#Forecasting models 
Eight models are utilized to test our forecasts, notably : 
1. Seasonal naive model 
2. Naive Model 
3. Arima model 
4. Stl+Ets model 
5. Arima + fourrier term model 
6. Tbats model 
7. NN model 
8. NN + fourrier term model

After testing these model, the best fit was utilized to fit the forecast. 
#Seasonal naive model
```{r}
seasonal_naive <- snaive(coal_ts_train, h=85) #the whole coal dataset
summary (seasonal_naive)
plot(seasonal_naive)
```

#Naive model

```{r}
naivem <- naive(coal_ts_train, h=85)
autoplot(naivem) +
        labs(title = "Coal Imports Naive forecast", 
             y = "Net Imports", 
             x = "Years")
```

## ARIMA model
```{r}
#coal forecast
arima_coal<- auto.arima(coal_ts_train)
arima_coal_for1 <- forecast(arima_coal, h=85)
autoplot(coal_ts_train, series="Original")+
  autolayer(arima_coal$fitted, series="ARIMA Model")+
  autolayer(arima_coal_for1$mean, series = "ARIMA Forecast")+
  labs(title = "Coal Imports Arima forecast",
       y = "Net Imports",
       x = "Years")
```

#STL + ETS model 

```{r}
ETS  <-  stlf(coal_ts_train, h=85)
autoplot(ETS) + ylab("Coal imports")
autoplot(coal_ts_train) +
  autolayer(ETS, series="STL + ETS",PI=FALSE) +
  labs(title = "Coal Imports ETS forecast",
       y = "Net Imports",
       x = "Years")
```

##ARIMA + FOURIER terms model

```{r ARIMA, echo=TRUE, message=FALSE, warning=FALSE}
ARIMA_Four_fit <- auto.arima(coal_ts_train, 
                             seasonal=FALSE,
                             lambda=0,
                             xreg=fourier(coal_ts_train,
                                          K=c(6))
                             )
# ARIMA Fourier
ARIMA_Four_model <- forecast(ARIMA_Four_fit,
                           xreg=fourier(coal_ts_train,
                                        K=c(6),
                                        h=85),
                           h=85
                           ) 

#Plot foresting results
autoplot(ARIMA_Four_model) + ylab("Coal Import")

#Plot model + observed data
autoplot(coal_ts_train) +
  autolayer(ARIMA_Four_model, series="ARIMA_FOURIER",PI=FALSE) +
  labs(title = "Coal Imports Arima Fourrier forecast",
       y = "Net Imports",
       x = "Years")
```

# TBATS Model

```{r TBATS, echo=TRUE, message=FALSE, warning=FALSE}

TBATS_fit <- tbats(coal_ts_train)
TBATS_for <- forecast(TBATS_fit, h=85)

#Plot foresting results
autoplot(TBATS_for)

#Plot model + observed data
autoplot(coal_ts_train) +
  autolayer(TBATS_for, series="TBATS",PI=FALSE)+
  labs(title = "Coal Imports TBATS forecast",
       y = "Net Imports",
       x = "Years")
```

# Neural Network Time Series Forecasts
```{r NNETAR, echo=TRUE, message=FALSE, warning=FALSE}
NN_fit <- nnetar(coal_ts_train, p=1, P=0, K=c(6))

NN_for <- forecast(NN_fit, h=85) 

#Plot foresting results
autoplot(NN_for)

#Plot model + observed data
autoplot(coal_ts_train) +
  autolayer(NN_for, series="Neural Network",PI=FALSE)+
  labs(title = "Coal Imports NN forecast",
       y = "Net Imports",
       x = "Years") 
```

# Neural Network +fourrier Forecasts
```{r NNETAR, echo=TRUE, message=FALSE, warning=FALSE}
NN_fit_reg <- nnetar(coal_ts_train, p=1,  P=0,
                 xreg=fourier(coal_ts_train, K=c(6)))

NN_for_reg <- forecast(NN_fit_reg, h=85,xreg=fourier(coal_ts_train, 
                                          K=c(6),h=85))

#Plot foresting results
autoplot(NN_for_reg)

#Plot model + observed data
 autoplot(coal_ts_train) +
  autolayer(NN_for_reg, series="Neural Network",PI=FALSE) +
  labs(title = "Coal Imports NN+regr forecast",
       y = "Net Imports",
       x = "Years")
```

## Checking accuracy of the eight models

```{r}
#Model 1: Seasonal Naive
Seasonaive_score <- accuracy (seasonal_naive$mean, coal_ts_test)

#Model 2 : Naive model 
Naive_score <- accuracy(naivem$mean, coal_ts_test)

#Model 3: ARIMA model 
ARIMA_score <- accuracy(arima_coal_for1$mean, coal_ts_test)

#Model 4 : ETS model 
ETS_scores <- accuracy(ETS$mean,coal_ts_test)  

#Model 5: ARIMA + Fourier 
ARIMA_Four_scores <- accuracy(ARIMA_Four_model$mean,coal_ts_test)
  
# Model 6:  TBATS 
TBATS_scores <- accuracy(TBATS_for$mean,coal_ts_test)

# Model 7:  Neural Network 
NN_scores <- accuracy(NN_for$mean,coal_ts_test)

# Model 8:  Neural Network 
NN_scores_reg <- accuracy(NN_for_reg$mean,coal_ts_test)
```

### Compare performance metrics

```{r}
#create data frame
scores <- as.data.frame(
  rbind(Seasonaive_score, Naive_score, ARIMA_score, ETS_scores, ARIMA_Four_scores, TBATS_scores, NN_scores, NN_scores_reg)
  )
row.names(scores) <- c("SEANAIV", "NAIVE", "ARMA", "ETS", "ARIMA+Fourier", "TBATS","NN", "NNREG")

#choose model with lowest RMSE
best_model_index <- which.min(scores[,"RMSE"])
cat("The best model by RMSE is:", row.names(scores[best_model_index,])) 
```

#Producting the accuracy table 

```{r echo=FALSE, message=FALSE, warning=FALSE}
kbl(scores, 
      caption = "Forecast Accuracy for Coal Import",
      digits = array(5,ncol(scores))) %>%
  kable_styling(full_width = FALSE, position = "center", latex_options = "hold_position") %>%
  #highlight model with lowest RMSE
  kable_styling(latex_options="striped", stripe_index = which.min(scores[,"RMSE"]))
```
Based on the forecasting models, the TBATS model produced the lowest RMSE (9.66603) and the MAPE (9.23292). Therefore, it is the best model according to our data; i.s it is likely to best suit our forecast for the 2030 period.

#Plotting the forecasted values 

```{r}
autoplot(coal_ts, PI=FALSE, series="Orignal") +
  autolayer(seasonal_naive, PI=FALSE, series="Seasonal Naive")+
  autolayer(naivem, PI = FALSE, series = "Naive") +
  autolayer(arima_coal_for1, PI=FALSE, series= "Arima")+
  autolayer(ETS, PI=FALSE, series="STL+ETS") +
  autolayer(ARIMA_Four_model, PI=FALSE, series="ARIMA + Fourier") +
  autolayer(TBATS_for,PI=FALSE, series="TBATS") +
  autolayer(NN_for,PI=FALSE, series="NN") +
  autolayer(NN_for_reg, PI=FALSE, series="NN fourrier")+
  xlab("Years") + ylab("Coal Import") +
  guides(colour=guide_legend(title="Forecasted"))
```
Reading the plot, the TBATS forecast line follows the historical trend more closely than others, with less deviations. This confirms the accuracy shown in the table as the TBATS model may accurately handle both the level and seasonality of the data.

# forcasting 2030 and produce values
```{r}
# Define the start and end date
start_date <- as.Date("2023-12-01")
end_date <- as.Date("2030-12-31")
dates <- seq.Date(start_date, end_date, by="month")

Best_model <- tbats(ts_faustin)
future_forecast <- forecast(Best_model, h=85)

forecasted_values <- future_forecast$mean
coalforecast <- data.frame(date = dates, load = round(forecasted_values, digits=0))
# Save the forecast dataframe to a CSV file
write.csv(coalforecast, file= "/Users/faustinkambale/Library/CloudStorage/OneDrive-DukeUniversity/Spring 2024 classes/Time Series 4 Energy/Competition/Faustin_Yilun/tsa-s24-competition/coalforecast.csv", row.names = FALSE)
```

#Ploting the forecast with best model
```{r}
autoplot(coal_ts) +
  autolayer(TBATS_for, series="TBATS",PI=FALSE)+
  autolayer(future_forecast, series = "Coal forecast", PI=FALSE) +
  ylab("Coal Import forecasting") 
```
This graph shows the forecasted values using the TBAT model on the historical data. The TBATS model captures the historical seasonality with the trend and provide accurate forecasts. The overall forecast suggest a decrease in coal imports. The TBAT model captures the historical pattern and forecasts the expected continued decrease in coal imports.
