---
title: "ENV797_TSA_project_S2024"
author: "Faustin Kambale Samantha Pace Drew Wolanski"
date: "2024-03-22"
output: pdf_document
editor_options: 
  chunk_output_type: console
---


```{r, message=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(cowplot)
library(sarima)
library(patchwork)
library(dplyr)
library(forecast)
library(readxl)
library(kableExtra)
```

## CO2 Emissions Forecasting : Predicting U.S. CO2 emissions from energy consumption

# Background & Research Question

1. Coal has been on the decline due to poor economics and increased scrutiny on emissions
2. Natural gas has concurrently been on the rise due to low prices caused by fracking revolution
   - Decrease in coal emissions partially offset by new natural gas plants
3. Project aims to forecast the CO2 emissions from coal, natural gas, and petroleum based on
historical data and determine if total CO2 emissions are declining as needed to meet climate goals
  - Biden goal to reduce GHG emissions by 50% below 2005 levels by 2030

# Data description

To verify our research question, we used data from the U.S. Energy Information Administration *Monthly Energy Review February 2024*

Data Category: Environment
Energy Sources: Coal, Natural Gas, Petroleum
Sampling frequency: Monthly
Training Data Date Range: Jan 1973-Nov 2023 (n=611)
Forecast Length: 6 years (2030)

## Loading the Dataset

```{r}
data <- read_excel("C:/Users/dreww/OneDrive/Documents/Homework/Time Series/TSA_Sp24/Project/Table_11.1_Carbon_Dioxide_Emissions_From_Energy_Consumption_by_Source.xlsx",  skip = 10)
```

##Data wrangling: remaning the variables

```{r}

#Inspecting the Data
data$Month <- as.Date(data$Month)

#Setting Data to work with
ourdata <- as.data.frame(data[,c("Month","coalimports", "naturalgas", "petroleumlessbioco2emissions")])
str(ourdata)
head(ourdata, 10)
tail(ourdata, 10)
summary(ourdata, 10)

ts_faustin <- ts(ourdata[,2], start=c(1973,1),frequency=12)
plot(ts_faustin)
ts_samantha <- ts(ourdata[,3], start=c(1973,1),frequency=12)
plot(ts_samantha)
ts_petrol <- ts(ourdata[,4], start=c(1973,1),frequency=12)
plot(ts_petrol)
```


```{r}
# Remove 2020
# Feb 2019-Feb 2020
petrol_2019 <- ts_petrol[555:566]
petrol_2021 <- ts_petrol[579:590]
covid_avg <- rowMeans(cbind(petrol_2019, petrol_2021))
ts_petrol_clean <- ts_petrol
ts_petrol_clean[567:578] <- covid_avg
```

##Data analysis 

# Ploting ACF and PACF

```{r}
faustin_acf <- Acf(ts_faustin, lag.max = 40)
faustin_pacf <-Pacf(ts_faustin, lag.max = 40)

plot_grid (
  autoplot(faustin_acf),
  autoplot(faustin_pacf)
)
```

# Decomposing Data

```{r}
faustin_decomp <- plot(decompose(ts_faustin, type = "additive"))
```

#Zooming in a small data sample 

```{r}
data_filtered <- filter(ourdata, year(Month) >= 2010)
ts_faustin1 <- ts(data_filtered[,2], start=c(2010,1),frequency=12)
plot(ts_faustin1)
faustin_decomp1 <- plot(decompose(ts_faustin1, type = "additive"))
```

In either model (short or long), we can notice a seasonal component in out dataset. 

##Forecasting 

```{r message=FALSE, warning=FALSE}
#create a subset data for training
n_for = 85
ts_faustin_train <- subset(ts_faustin,
                          end = length(ts_faustin)-n_for)

#create a subset data for experiment
ts_faustin_test <- subset(ts_faustin,
                      start = length(ts_faustin)-n_for+1, end = length(ts_faustin))

autoplot(ts_faustin_train)
autoplot(ts_faustin_test)
```
# Training and test sets
```{r}
n_train <- round(0.8*length(ts_petrol))
n_test <- length(ts_petrol)-n_train
petrol_train <- ts_petrol_clean[1:n_train]
petrol_test <- ts_petrol_clean[(n_train+1):length(ts_petrol)]

```

#Forecasting models 
#Seasonal naive model
```{r}
seasonal_naive <- snaive(ts_faustin_train, h=12)
summary (seasonal_naive)
plot(seasonal_naive)
```

#Naive model

```{r}
naivem <- naive(ts_faustin_train,h=12)
summary (naivem)
plot(naive)
```

## ARIMA model
```{r}
arima_whole<- auto.arima(ts_faustin)
arima<- auto.arima(ts_faustin_train)
arima_for <- forecast(arima, h=85)

autoplot(ts_faustin, series="Original")+
  autolayer(arima_whole$fitted, series="ARIMA Model")+
  autolayer(arima_for1$mean, series = "ARIMA Forecast")
```
# Petroleum ARIMA
```{r}
petrol_arima_train <- auto.arima(petrol_train)
petrol_arima_testfor <- forecast(petrol_arima_train,h=n_test)
petrol_arima_acc <- accuracy(petrol_arima_testfor,petrol_test)
petrol_arima_fit <- auto.arima(ts_petrol_clean)
petrol_arima_for <- forecast(petrol_arima_fit,h=85)
```
# Petroleum TBATS
```{r}
petrol_tbats_train <- tbats(petrol_train)
petrol_tbats_testfor <- forecast(petrol_tbats_train,h=n_test)
petrol_tbats_acc <- accuracy(petrol_tbats_testfor,petrol_test)
petrol_tbats_fit <- tbats(ts_petrol_clean)
petrol_tbats_for <- forecast(petrol_tbats_fit,h=85)
```

# Simple NN

```{r}
# Fit simple (p=24,P=12) NN model
petrol_nn_train_11 <- nnetar(petrol_train,p=24,P=12)
petrol_nn_testfor_11 <- forecast(petrol_nn_train_11,h=n_test)
petrol_nn_acc_11 <- accuracy(petrol_nn_testfor_11,petrol_test)
cat("Neural Network: p = 1, P = 1, Training Accuracy (MAPE):", petrol_nn_acc_11[9],", Test Accuracy (MAPE):", petrol_nn_acc_11[10])

# Train on whole data set and forecast to 2030
petrol_nn_11_fit <- nnetar(ts_petrol_clean,p=24,P=12)
petrol_nn_11_for <- forecast(petrol_nn_11_fit, h=85)
```

# Petroleum NN Optimization
```{r}
petrol_nn_acc_min <- 10
p_opt <- 0
P_opt <-20
for (p in seq(0:15)){
  print(p)
  for (P in seq(0:15)){
    petrol_nn_train <- nnetar(petrol_train,p=p,P=P)
    petrol_nn_testfor <- forecast(petrol_nn_train,h=n_test)
    petrol_nn_acc <- accuracy(petrol_nn_testfor,petrol_test)
    if ((petrol_nn_acc[9] + petrol_nn_acc[10]) < petrol_nn_acc_min){
      petrol_nn_acc_min <- petrol_nn_acc[9] + petrol_nn_acc[10]
      p_opt <- p
      P_opt <- P
    }
  }
}
print(petrol_nn_acc_min, p_opt, P_opt)

```

# Optimized Petroleum NN
```{r}
# Fit optimized model
petrol_nn_train_opt <- nnetar(petrol_train,p=p_opt,P=P_opt)
petrol_nn_testfor_opt <- forecast(petrol_nn_train_opt,h=n_test)
petrol_nn_acc_opt <- accuracy(petrol_nn_testfor_opt,petrol_test)
cat("Optimized Neural Network: p =",p_opt, ", P =",P_opt,", Training Accuracy (MAPE):", petrol_nn_acc_opt[9],", Test Accuracy (MAPE):", petrol_nn_acc_opt[10])

# Train on whole data set and forecast to 2030
petrol_nn_fit <- nnetar(ts_petrol_clean,p=p_opt,P=P_opt)
petrol_nn_for <- forecast(petrol_nn_fit, h=85)
```

# Plot Forecasts
```{r}

autoplot(ts_petrol,series="Petroleum Time Series History") +
  autolayer(petrol_nn_fit$fitted,series = "Petroleum NN Model (p = 13, P = 4)") +
  autolayer(petrol_nn_for$mean, series = "Petroleum Optimized NN Forecast") +
  #Add other forecasts to same plot
  autolayer(petrol_nn_11_for$mean, series = "Petroleum Simple NN Forecast") +
  autolayer(petrol_arima_for$mean, series = "Petroleum ARIMA Forecast") +
  autolayer(petrol_tbats_for$mean, series = "Petroleum TBATS Forecast") +
  
  xlab("Year") + ylab("Petroleum CO2 Emissions (Million Metric Tons CO2)") +
  ggtitle("Petroleum CO2 Emissions 2030 Forecast") 
```


```{r}
autoplot(ts_petrol, series="Petroleum Time Series History") +
  autolayer(petroleum_forecast$mean, series = "Petroleum ARIMA Forecast") +
  autolayer(petroleum_arima$fitted, series="Petroleum ARIMA Model")+
  #scale_color_manual(values = c("purple", "red", "blue")) +
  xlab("Year") + ylab("Petroleum CO2 Emissions (Million Metric Tons CO2)") +
  ggtitle("Petroleum CO2 Emissions 2030 Forecast") 
  # + theme_minimal()
```

#STL + ETS model 

```{r}
ETS  <-  stlf(ts_faustin_train,h=12)
autoplot(ETS) + ylab("Coal imports")

#Plot model + observed data
autoplot(ts_faustin_train) +
  autolayer(ETS, series="STL + ETS",PI=FALSE) +
  ylab("Coal imports")
```

##ARIMA + FOURIER terms model

```{r ARIMA, echo=TRUE, message=FALSE, warning=FALSE}
ARIMA_Four_fit <- auto.arima(ts_faustin_train, 
                             seasonal=FALSE,
                             lambda=0,
                             xreg=fourier(ts_faustin_train,
                                          K=c(6))
                             )
# ARIMA Fourier
ARIMA_Four_model <- forecast(ARIMA_Four_fit,
                           xreg=fourier(ts_faustin_train,
                                        K=c(6),
                                        h=12),
                           h=12
                           ) 

#Plot foresting results
autoplot(ARIMA_Four_model) + ylab("Power Demand")

#Plot model + observed data
autoplot(ts_faustin_train) +
  autolayer(ARIMA_Four_model, series="ARIMA_FOURIER",PI=FALSE) +
  ylab("Coal Import")
```

# TBATS Model

```{r TBATS, echo=TRUE, message=FALSE, warning=FALSE}

TBATS_fit <- tbats(ts_faustin_train)
TBATS_for <- forecast(TBATS_fit, h=12)

#Plot foresting results
autoplot(TBATS_for) +
  ylab("Coal Import") 

#Plot model + observed data
autoplot(ts_faustin_train) +
  autolayer(TBATS_for, series="TBATS",PI=FALSE)+
  ylab("Coal Import") 
```

# Neural Network Time Series Forecasts
```{r NNETAR, echo=TRUE, message=FALSE, warning=FALSE}
NN_fit <- nnetar(ts_faustin_train,
                 p=1,
                 P=0,
                 K=c(6))

NN_for <- forecast(NN_fit, h=12) 

#Plot foresting results
autoplot(NN_for) +
  ylab("Coal Import") 

#Plot model + observed data
autoplot(ts_faustin_train) +
  autolayer(NN_for, series="Neural Network",PI=FALSE)+
  ylab("Coal Import") 
```

# Model 5: Neural Network Time Series Forecasts
```{r NNETAR, echo=TRUE, message=FALSE, warning=FALSE}
NN_fit_reg <- nnetar(ts_faustin_train,
                 p=1,
                 P=0,
                 xreg=fourier(ts_faustin_train, K=c(6)))

NN_for_reg <- forecast(NN_fit_reg, h=12,xreg=fourier(ts_faustin_train, 
                                          K=c(6),h=12))

#Plot foresting results
autoplot(NN_for_reg) +
  ylab("Coal Import") 

#Plot model + observed data
autoplot(ts_faustin_train) +
  autolayer(NN_for_reg, series="",PI=FALSE)+
  ylab("Coal Import") 
```

## Checking accuracy of the four models

```{r}
#we can define the cutt off from here

#Model 1: Seasonal Naive
Seasonaive_score <- accuracy (seasonal_naive$mean, ts_faustin_test)

#Model 2 : Naive model 
Naive_score <- accuracy(naivem$mean, ts_faustin_test)

#Model 3: ARIMA model 
ARIMA_score <- accuracy(arima_for$mean, ts_faustin_test)

#Model 4 : ETS model 
ETS_scores <- accuracy(ETS$mean,ts_faustin_test)  

#Model 5: ARIMA + Fourier 
ARIMA_Four_scores <- accuracy(ARIMA_Four_model$mean,ts_faustin_test)
  
# Model 6:  TBATS 
TBATS_scores <- accuracy(TBATS_for$mean,ts_faustin_test)

# Model 5:  Neural Network 
NN_scores <- accuracy(NN_for$mean,ts_faustin_test)

# Model 5:  Neural Network 
NN_scores_reg <- accuracy(NN_for_reg$mean,ts_faustin_test)
```

### Compare performance metrics

```{r}
#create data frame
scores <- as.data.frame(
  rbind(Seasonaive_score, Naive_score, ARIMA_score, ETS_scores, ARIMA_Four_scores, TBATS_scores, NN_scores, NN_scores_reg)
  )
row.names(scores) <- c("SEANAIV", "NAIVE", "ARMA", "ETS", "ARIMA+Fourier", "TBATS","NN", "NNREG")

#choose model with lowest RMSE
best_model_index <- which.min(scores[,"RMSE"])
cat("The best model by RMSE is:", row.names(scores[best_model_index,])) 
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
kbl(scores, 
      caption = "Forecast Accuracy for Coal Import",
      digits = array(5,ncol(scores))) %>%
  kable_styling(full_width = FALSE, position = "center", latex_options = "hold_position") %>%
  #highlight model with lowest RMSE
  kable_styling(latex_options="striped", stripe_index = which.min(scores[,"RMSE"]))
```
